Supplement to Automatic Matching of Bullet Lands

### 2015-12-23 imaging-paper.Rnw
more into supplement

\documentclass[aoas]{imsart}

@@

In 2009, the National Academy of Sciences published a report questioning the scientific validity of many forensic methods including firearm examination. ~\citep{NAS:2009}. The report states that ``[m]uch forensic evidence -- including, for example, bite marks and firearm and toolmark identification is introduced in criminal trials without any meaningful scientific validation, determination of error rates, or reliability testing to explain the limits of the discipline."

@@

\section{Methodology}
Throughout this paper, we work with images from the James Hamby Consecutively Rifled Ruger Barrel Study~\citep{hamby:2009}. Ten consecutively rifled Ruger P-85 pistol barrels were obtained from the manufacturer and fired to produce known test bullets and unknown bullets for comparison.

@@

\hh{In an approach to analyze the striation pattern we extract a {\emph bullet profile} or {\emph signature} \cite{ma:2004} by taking a cross section across the surface measurements for a fixed value of $x$.}
%An initial naive approach to analyzing the striation patterns leads us to fixing a particular value of a coordinate, and then producing a plot of the surface measurements across. 
Figure \ref{fig:fixedX} shows a plot of a bullet signature. It can be seen that the global structure of the land dominates the appearance of the plot. The grooves can be clearly identified on the left and right side, and the curvature of the surface is the most visible feature in the middle.

@@

The smooth curve on the plot represents a segment of a perfect circle with the same radius as the bullet. \hh{While the circle is an obvious first choice for fitting the structure, it does not completely capture the bullet surface after it was fired. A discussion of the fit and the remaining residual structure can be found in the supplement.} \hh{XXX Reference to the supplement? XXX}
%It is clear that our next approach must be to model the overall structure in order to focus on the deviations, given that the deviations should correspond to bullet markings.

%\subsection{Loess Fit}
\hh{Instead, we use multiple loess fits to model the overall structure in order to identify the bullet markings. }

\subsection{Identify Grooves}
%In an attempt to improve upon the previous algorithm, w
We first have to \hh{identify} the \hh{locations of the left and right}  groove in the image. \hh{The grooves are not assumed to contain information relevant for determining matches, they also dominate the structure, and therefore need  to be removed.}  
Fortunately, the location and appearance of the grooves in the surface profiles is quite consistent.
Surface measurements reach local maxima around the peak of the groove at either end of the range of $y$, we can then follow the descent of the surface measurements inwards to the valley of the groove. 
This marks the point at which we should trim the image. The procedure can be described as follows:

\begin{enumerate}
    \item Fix an $x$ value \hh{to extract a bullet signature} (Figure~\ref{fig:loess_step1}, with $x$ set to 243.75$\mu m$).
    \item For each $y$ value, smooth out any deviations occurring near the minima by twice applying a rolling average with a pre-set \emph{smoothing factor} $s$. (Figure~\ref{fig:loess_step3}, smoothing factor $s = 35$ corresponding to 55$\mu m$).
%    \item For each smoothed $y$ value, compute another rolling average using the same smoothing factor as above. (Figure~\ref{fig:loess_step3}).
    \item Determine the location of the peak of the left groove by finding the first double-smoothed value $y_i$ that is the maximum within its smoothing window (e.g.\ such that $y_i > y_{i - 1}$ and $y_i > y_{i + 1}$, where $i$ is between 1  and $\lfloor s/2 \rfloor$). We call the location of this peak $p_{\ell}$ (see Figure~\ref{fig:loess_step47}). 
    \item Similarly, determine the location of the valley of the left groove by finding the first double-smoothed $y_j$ that is the minimum within its smoothing window. Call the location of this valley $v_{\ell}$.
    \item Reverse the order of the $y$ values and repeat the previous two steps to find the peak and valley of the right groove, $(p_{r}, v_{r})$.
    \item Trim the surface measurements to values within the two grooves (i.e.\ remove all records with $y_i < v_{\ell}$ and $y_i > v_{r}$) (see Figure~\ref{fig:loess_step47}).
\end{enumerate}

@@

Next, we fit a loess regression to the data. Figure~\ref{fig:loess_fit} provides a look at the loess fit, in blue, overlaid against the processed image of part 1 of bullet 1. The fit seems to do a reasonable job of capturing the structure of the image. Figure~\ref{fig:loess_resid} shows the residuals from this fit. We can reproduce a similar figure to the one from the circular fits by overlaying the residuals from each loess fit for bullet 2 onto an image of the loess residuals from bullet 1-5. This is shown in Figure~\ref{fig:loessfits}. Bullet 2-1 is still the best match for bullet 1-5, but now the details of the match are more pronounced.

\begin{figure}[hbtp]
  \centering
\begin{subfigure}[t]{.49\textwidth}\centering
\caption{\label{fig:loess_fit} Loess fit for bullet 1-5.}
<<loess_fit, echo=FALSE, fig.width=10, fig.height=3.5, out.width='\\textwidth', warning=FALSE,  message=FALSE>>=

@@

my.loess$fitted + ylab("Surface measurements")
@
\end{subfigure}
\begin{subfigure}[t]{.49\textwidth}\centering
\caption{\label{fig:loess_resid} Residuals of loess fit for bullet 1-5.}
<<loess_resid, echo=FALSE, fig.width=10, fig.height=3.5, out.width='\\textwidth', warning=FALSE>>=
my.loess$resid + ylab("Residuals of loess Fit")
@
\end{subfigure}
\caption{Fit and residuals of a loess fit to bullet 1-5 (Barrel 1).}
\end{figure}

@@

actual.df <- my.loessactual$data[1:minrows,]
melt.df$land <- melt.df$part
qplot(y, resid, data=melt.df, #colour=factor(x), 
      geom="line", size=I(.75), colour=I("grey70")) + 
  facet_wrap(~land, ncol=3, labeller="label_both") + 
  scale_colour_brewer(palette="Paired") +
  theme_bw() + theme(legend.position="bottom") + 
  geom_line(data = actual.df, aes(y, resid), colour="black", size=.75, alpha=0.5) + ggtitle("Bullet 1-5 in black") + ylab("Residuals of loess Fit")
@    
\caption{\label{fig:loessfits} Residuals of a loess fit to bullets 1 and 2 from barrel 1. Bullet 1-5 is shown in black. }
\end{figure}

% \section{Web Application}
% We have implemented a web interface to automate many of the steps outlined in the previous section. The web interface, developed using Shiny~\citep{shiny}, allows a user to upload two x3p files representing scans of a bullet between two grooves. A 3D viewer, using Plotly~\citep{plotly}, allows the user to rotate and zoom to examine surface features of the images. A screenshot of the application is available in Figure \ref{fig:app}. The application is available at the url \url{http://erichare.shinyapps.io/x3prproto}.
% 
% \begin{figure*}
% \includegraphics[width=\textwidth]{images/app.png}
% \caption{\label{fig:app}{Screenshot of the web interface for analyzing bullet images.}}
% \end{figure*}
% 
% The mouse is the primary control input for interacting with the surface plots. One can mouse over the plots to view the coordinate values for each point. Clicking and dragging anywhere on the plot will allow rotation. Scrolling up and down allows the user to zoom out and and respectively. There are also a set of controls available at the top right of the page, which will allow the selection of some other rotation methods, taking of screenshots, and reseting the camera to the default view.
% 
% Several lighting options are available to tweak in order to highlight markings in on the bullets more strongly:
% 
% \begin{enumerate}
%     \item \textbf{Ambient Lighting} - Lighting which is used to simulate the light naturally present in the environment.
%     \item \textbf{Diffuse Lighting} - Setting this value higher will cause the surface to re-emit a higher proportion of light.
%     \item \textbf{Specular Lighting} - The lighting which appears on the surface when illuminated.
%     \item \textbf{Roughness Lighting} - The amount of light reflected by the roughness of the surface.
%     \item \textbf{Fresnel Lighting} - Higher values create a wider, softer lighting across the surface.
% \end{enumerate}
% 
% The application includes a button which allows the automatic creation of a residual plot from the loess fits of two bullets. An x coordinate value can be selected as a cross cut, and cross sections of the two bullets at the x value are taken. The previously described loess regression procedure is applied to the two cross sections. A residual plot is produced beneath the surface plots allowing for an assessment of the similarity of the markings. Users need to perform this procedure at different x values in order to see whether the markings indicate a match at different points along the bullet.
% 
% Enhancements to the web application are still in progress. In particular, much of the automatic matching functionality described in the next section will be added in the future.

\section{Automatic matching}
Applying the loess fit to a range of different surface cross cuts (see Figure~\ref{fig:manualmatch-rgl}, with cross cuts from 50 to 150) shows the striation marks from two bullets. Cross cuts of bullet 1 are shown on the left (cross cuts with values below 100) and cross cuts of bullets 2 are shown on the right (cross cuts with values above 100). Many of the striation marks align, allowing for an easy visual assessment of the shared similarities between the two bullets.

@@

lines <- matches %>% group_by(lineid) %>% summarise(
  meany = mean(y, na.rm=T),
  miny = min(y, na.rm=T),
  maxy = max(y, na.rm=T) + 1.5625,
  match = isMatch(lineid, type),
  type = type1[1]
)
lines <- subset(lines, lineid != 0)
@
\begin{figure}[hbtp]
\centering
\begin{subfigure}[b]{.49\textwidth}\centering
\caption{Cross section at $x = 100$.\label{fig:crosscut}}{%
<<crosscuts, echo=FALSE, dependson='twolands100', fig.width=7, fig.height=3, out.width='\\textwidth'>>=
qplot(y, resid, group=bullet, colour=bullet, data=lof, 
      geom="line") + 
  scale_colour_manual("", values=cols) + theme_bw() +
  theme(legend.position = c(.9, .2),
        legend.background = element_rect(fill=alpha('white', 0.4))) + 
  ylab("Residuals of loess fit")
@
    }
\end{subfigure}    
\begin{subfigure}[b]{.49\textwidth}\centering
    \caption{Adjusted (bullet 1-5 horizontally shifted) cross sections at $x = 100$.\label{fig:crosscutX}}{%
<<crosscutsX, echo=FALSE, dependson='twolands100', fig.width=7, fig.height=3, out.width='\\textwidth'>>=
# qplot(y, resid, group=bullet, colour=bullet, data=lofX, 
#       geom="line", alpha=I(0.6)) + 
#   scale_colour_brewer(palette="Set1") + theme_bw() +
#   theme(legend.position="none")
qplot(y+min(lof$y), resid, group=bullet, colour=bullet, data=lofY, 
      geom="line") + xlab("y") +
  scale_colour_manual("", values=cols) + theme_bw() +
  theme(legend.position="none") + ylab("Residuals of loess fit")
@
    }
\end{subfigure}
\caption{\label{fig:cross100}Lines showing surface cross sections at $x = 100$. A horizontal shift of the values of bullet 1-5 to the right shows the similarity of the striation marks.}
\end{figure}

\subsection{Horizontal Alignment}
Using (for now) a single cross section across the surface, we extract the values and remove the class characteristics using a loess fit as before. This enables us to focus on the individual characterists. Striation marks show up in this representations as peaks and valleys. Figure~\ref{fig:cross100} shows lines of the surface cross section at $x = 100$. A horizontal shift of one of the lines (result shown in (b)) emphasizes the strong similarities.
This horizontal shift is based on the cross-correlation between the two surface cross sections. Let $f(t)$ and $g(t)$ define the values of the cross section at $t$, then the cross-correlation between $f$ and $g$ at lag $k$ is defined as
\[
(f * g) (k) = \sum_t f(t+k) g(t),
\]
with suitably defined limits for the summation. % (because variances of $f$ and $g$ do not change, the above equation is proportional to the correlation).

\begin{figure}[hbtp]
  \centering
<<ccf, echo=FALSE, fig.width = 7, fig.height = 3, out.width = '.5\\textwidth'>>=
lag <- ccf$lag[which.max(ccf$acf)]
qplot(x=ccf$lag, xend=ccf$lag, y = 0, yend= ccf$acf, geom="segment", colour = I("grey50")) + 
  theme_bw() + ylab("Correlation") + xlab("Lag k") +
  geom_segment(aes(x = lag, xend = lag, y = 0, yend =  max(ccf$acf)), colour="black") + 
  scale_x_continuous(breaks = c(-100, -50, lag, 0, 50, 100),
                     minor_breaks = c(-75,-25,25, 75))
@
\caption{\label{fig:ccf}Cross-correlation between the two cross sections shown in Figure~\ref{fig:crosscut} at lags between -100 and 100. At a lag of -10 the correlation peaks, indicating the largest amount of agreement between the cross sections. Figure~\ref{fig:crosscutX} shows the lag-shifted cross sections.}
\end{figure}

\subsection{Varying cross sections}
The cross section at which a comparison is done matters -- as cross sections are further apart, more pronounced differences between the cross sections show up. This poses both a caveat to matching attempts as well as an opportunity \hh{for quality control}: on the one hand we have to be aware of which cross section level was used in a matching, but on the other hand we can use multiple cross sections of the same bullet for an initial assessment of its quality. Both of these situations are assessed and discussed in more detail below: 
Figure~\ref{fig:crosscuts} shows a sequence of cross sections for bullet 1-5 (barrel 1) that are taken 50$\mu m$ apart, between 150$\mu m$  and 400$\mu m$. These are compared to a cross section at 100$\mu m$. Initially this comparison constitutes an almost perfect match between the two cross sections. However, the match quickly deteriorates with increasing distance to non-matches at a cross section of $x = 300\mu m$ or higher.  Only if the cross section is within 150$\mu m$ do we get a good visual match even when we know that the same bullet surface is being used. 
Given that we have to expect some variation in nominally the same cross section levels due to manual alignments in microscopes, we should take cross section levels into account in the automatic matching routine by evaluating matches at several cross sections. 
\begin{figure}[hbtp]
  \centering
<<crosscuts-vary, echo=FALSE, fig.width = 12, fig.height = 7, out.width = '\\textwidth', cache=TRUE, warning = FALSE>>=
paths <- file.path(datadir, dir(datadir))
crosscuts <- seq(100, 400, by = 50)
lof <- processBullets(paths[5], x = crosscuts)
lof$bullet <- paste(lof$bullet, lof$x)

reslist <- lapply(crosscuts[-1], function(cc) {
#  browser()
  b2 <- subset(lof, x %in% c(cc, 100))
  lofX <- bulletSmooth(b2)
  bAlign = bulletAlign(lofX)
  lofX <- bAlign$bullet
    b12 <- unique(b2$bullet)
  peaks1 <- get_peaks(subset(lofX, bullet==b12[1]), smoothfactor = 25)
  peaks2 <- get_peaks(subset(lofX, bullet == b12[2]), smoothfactor = 25)

lop <- lapply(reslist, function(res) {
ggplot() +
  theme_bw() + 
  geom_rect(aes(xmin=xmin, xmax=xmax, fill=factor(type)), show.legend=FALSE, ymin=-6, ymax=5, data=res$lines, alpha=0.2) +
  geom_line(aes(x = y, y = l30, linetype=bullet),   data = res$bullets, alpha=0.6) +
  scale_colour_brewer("", palette="Set1", na.value=alpha("grey50", alpha=0.5)) +
  scale_linetype_discrete("") +
  scale_fill_brewer("", palette="Set2", na.value=alpha("grey50", alpha=0.5)) +
  theme(legend.position = c(1,1.2), legend.justification=c(1,1),
        legend.background = element_rect(fill=alpha('white', 0.4))) + 
  ylim(c(-6,6)) +
  geom_text(aes(x = meany), y= -5.5, label= "x", data = subset(res$lines, !match)) +
  geom_text(aes(x = meany), y= -5.5, label= "o", data = subset(res$lines, match)) +
    ylab("") + xlab("")
})

grid.arrange(lop[[1]], lop[[2]], lop[[3]], lop[[4]], lop[[5]], lop[[6]],
             ncol = 2)
@
\caption{\label{fig:crosscuts}Overview of the variations in the surface along different cross sections. Cross section at $x = 100$ is compared to cross sections every 50$\mu m$. With every step away from the original cross section site, the number of differences between the cross sections increases, and the number of maximum CMS decreases from initially 14 to  3 or fewer at $x = 300\mu m$ and higher. }
\end{figure}
%\afterpage{\clearpage}

@@

\begin{figure}[hbtp]
  \centering
<<tree:forest, echo=FALSE, fig.width=6, fig.height=6.25, out.width='.5\\textwidth', warning = FALSE>>=
qplot(pred, forest, geom="jitter", data=bstats, colour=match, 
      shape=match) + theme_bw() +
  theme(legend.position="bottom") +
  geom_vline(xintercept=0.5, colour="grey50") +
  geom_hline(yintercept=0.5, colour="grey50") + 
  xlab("Predictions from decision tree") +
  ylab("Predictions from random forest")   +
  scale_colour_brewer(palette="Paired", labels=c("KNM", "KM")) +
  scale_shape_discrete(labels=c("KNM", "KM"))
@
\caption{\label{fig:tree-forest}Scatterplot of prediction results from the tree (horizontal) and the forest (vertical). Each point corresponds to a land-to-land comparison. Known matches are indicated in blue. Matches predicted by the decision tree are on the right half of the plot, matches predicted by the random forest are in the upper half. }
\end{figure}

>>>>

\documentclass[aoas, preprint]{imsart}

@@

In 2009, the National Academy of Sciences published a report~\citep{NAS:2009} questioning the scientific validity of many forensic methods including firearm examination. The report states that ``[m]uch forensic evidence -- including, for example, bite marks and firearm and toolmark identification is introduced in criminal trials without any meaningful scientific validation, determination of error rates, or reliability testing to explain the limits of the discipline."

@@

\section{Bullet signatures}
Throughout this paper, we work with images from the James Hamby Consecutively Rifled Ruger Barrel Study~\citep{hamby:2009}. Ten consecutively rifled Ruger P-85 pistol barrels were obtained from the manufacturer and fired to produce known test bullets and unknown bullets for comparison.

@@

\hh{In an approach to analyze the striation pattern we extract a {\emph bullet profile} \cite{ma:2004} by taking a cross section across the surface measurements for a fixed value of $x$.}
%An initial naive approach to analyzing the striation patterns leads us to fixing a particular value of a coordinate, and then producing a plot of the surface measurements across. 
Figure \ref{fig:fixedX} shows a plot of the side profile of a bullet land. It can be seen that the global structure of the land dominates the appearance of the plot. The grooves can be clearly identified on the left and right side, and the curvature of the surface is the most visible feature in the middle.

@@

The smooth curve on the plot represents a segment of a perfect circle with the same radius as the bullet. \hh{While the circle is an obvious first choice for fitting the structure, it does not completely capture the bullet surface after it was fired. A discussion of a circular fit and the remaining residual structure can be found in the supplement.} \hh{XXX Reference to the supplement? XXX}
%It is clear that our next approach must be to model the overall structure in order to focus on the deviations, given that the deviations should correspond to bullet markings.

%\subsection{Loess Fit}
\hh{Instead of a circular fit, we use multiple loess fits to model the overall structure in order to identify the bullet markings. }

\subsection{Identifying grooves}
%In an attempt to improve upon the previous algorithm, w
We first have to \hh{identify} the \hh{locations of the left and right}  groove in the image. \hh{The grooves are not assumed to contain information relevant for determining matches, they also dominate the structure, and therefore need  to be removed.}  
Fortunately, the location and appearance of the grooves in the surface profiles is quite consistent.
Surface measurements reach local maxima around the peak of the groove at either end of the range of $y$, we can then follow the descent of the surface measurements inwards to the valley of the groove. 
The location of the valleys  mark the points at which we should trim the image. The procedure can be described as follows:

\begin{enumerate}
    \item Fix an $x$ value \hh{to extract a bullet profile} (Figure~\ref{fig:loess_step1}, with $x$ set to 243.75$\mu m$).
    \item For each $y$ value, smooth out any deviations occurring near the minima by twice applying a rolling average with a pre-set \emph{smoothing factor} $s$. (Figure~\ref{fig:loess_step3}, smoothing factor $s = 35$ corresponding to 55$\mu m$).
%    \item For each smoothed $y$ value, compute another rolling average using the same smoothing factor as above. (Figure~\ref{fig:loess_step3}).
    \item Determine the location of the peak of the left groove by finding the first double-smoothed value $y_i$ that is the maximum within its smoothing window (e.g.\ such that $y_i > y_{i - 1}$ and $y_i > y_{i + 1}$, where $i$ is between 1  and $\lfloor s/2 \rfloor$). We call the location of this peak $p_{\ell}$ (see Figure~\ref{fig:loess_step47}). 
    \item Similarly, determine the location of the valley of the left groove by finding the first double-smoothed $y_j$ that is the minimum within its smoothing window. Call the location of this valley $v_{\ell}$.
    \item Reverse the order of the $y$ values and repeat the previous two steps to find the peak and valley of the right groove, $(p_{r}, v_{r})$.
    \item Trim the surface measurements to values within the two grooves (i.e.\ remove all records with $y_i < v_{\ell}$ and $y_i > v_{r}$) (see Figure~\ref{fig:loess_step47}).
\end{enumerate}

@@

\subsection{Removing curvature}
Next, we fit a loess regression to the data.

\hh{XXX Introduce Loess formula with all of its details in here. we are using the default of {\tt loess}, which has a span of 0.75 and uses a polynomial of degree 2. XXX}
Figure~\ref{fig:loess_fit} provides a look at the loess fit, in blue, overlaid against the processed image of bullet 1-5. The fit seems to do a reasonable job of capturing the structure of the image. Figure~\ref{fig:loess_resid} shows the residuals from this fit. These residuals are called the \emph{signature of bullet 1-5.}
%
\begin{figure}[hbtp]
  \centering
\begin{subfigure}[t]{.49\textwidth}\centering
\caption{\label{fig:loess_fit} Loess fit for bullet 1-5.}
<<loess_fit, echo=FALSE, fig.width=10, fig.height=3.5, out.width='\\textwidth', warning=FALSE,  message=FALSE>>=

@@

my.loess$fitted + ylab("Surface measurements")
@
\end{subfigure}
\begin{subfigure}[t]{.49\textwidth}\centering
\caption{\label{fig:loess_resid} Residuals of loess fit for bullet 1-5.}
<<loess_resid, echo=FALSE, fig.width=10, fig.height=3.5, out.width='\\textwidth', warning=FALSE>>=
my.loess$resid + ylab("Residuals of loess Fit")
@
\end{subfigure}
\caption{Fit and residuals of a loess fit to bullet 1-5 (Barrel 1).}
\end{figure}
%
Figure~\ref{fig:loessfits} shows the signatures of all lands of bullet 2 (in grey) fired from the same barrel as bullet 1. The darker line overlaid on each of the signatures, shows signature from bullet 1-5. 
%We can reproduce a similar figure to the one from the circular fits by overlaying the residuals from each loess fit for bullet 2 onto an image of the loess residuals from bullet 1-5. This is shown in Figure~\ref{fig:loessfits}. 
Bullet 2-1 is the best match for bullet 1-5, but the alignment is not quite  correct.

@@

actual.df <- my.loessactual$data[1:minrows,]
melt.df$land <- melt.df$part
qplot(y, resid, data=melt.df, #colour=factor(x), 
      geom="line", size=I(.75), colour=I("grey70")) + 
  facet_wrap(~land, ncol=3, labeller="label_both") + 
  scale_colour_brewer(palette="Paired") +
  theme_bw() + theme(legend.position="bottom") + 
  geom_line(data = actual.df, aes(y, resid), colour="black", size=.75, alpha=0.5) + ggtitle("Signatures of Bullet 1 in grey, Bullet 1-5 in black") + ylab("Residuals of loess fit")
@    
\caption{\label{fig:loessfits} Residuals of a loess fit to bullets 1 and 2 from barrel 1. Bullet 1-5 is shown in black. }
\end{figure}

% \section{Web Application}
% We have implemented a web interface to automate many of the steps outlined in the previous section. The web interface, developed using Shiny~\citep{shiny}, allows a user to upload two x3p files representing scans of a bullet between two grooves. A 3D viewer, using Plotly~\citep{plotly}, allows the user to rotate and zoom to examine surface features of the images. A screenshot of the application is available in Figure \ref{fig:app}. The application is available at the url \url{http://erichare.shinyapps.io/x3prproto}.
% 
% \begin{figure*}
% \includegraphics[width=\textwidth]{images/app.png}
% \caption{\label{fig:app}{Screenshot of the web interface for analyzing bullet images.}}
% \end{figure*}
% 
% The mouse is the primary control input for interacting with the surface plots. One can mouse over the plots to view the coordinate values for each point. Clicking and dragging anywhere on the plot will allow rotation. Scrolling up and down allows the user to zoom out and and respectively. There are also a set of controls available at the top right of the page, which will allow the selection of some other rotation methods, taking of screenshots, and reseting the camera to the default view.
% 
% Several lighting options are available to tweak in order to highlight markings in on the bullets more strongly:
% 
% \begin{enumerate}
%     \item \textbf{Ambient Lighting} - Lighting which is used to simulate the light naturally present in the environment.
%     \item \textbf{Diffuse Lighting} - Setting this value higher will cause the surface to re-emit a higher proportion of light.
%     \item \textbf{Specular Lighting} - The lighting which appears on the surface when illuminated.
%     \item \textbf{Roughness Lighting} - The amount of light reflected by the roughness of the surface.
%     \item \textbf{Fresnel Lighting} - Higher values create a wider, softer lighting across the surface.
% \end{enumerate}
% 
% The application includes a button which allows the automatic creation of a residual plot from the loess fits of two bullets. An x coordinate value can be selected as a cross section, and cross sections of the two bullets at the x value are taken. The previously described loess regression procedure is applied to the two cross sections. A residual plot is produced beneath the surface plots allowing for an assessment of the similarity of the markings. Users need to perform this procedure at different x values in order to see whether the markings indicate a match at different points along the bullet.
% 
% Enhancements to the web application are still in progress. In particular, much of the automatic matching functionality described in the next section will be added in the future.

\section{Automatic matching}
Applying the loess fit to a range of different surface cross sections (see Figure~\ref{fig:manualmatch-rgl} for cross sections from 50$\mu m$ to 150 $\mu m$) shows the 3d striation marks from two bullets. Signatures of bullet 1 are shown on the left (all cross sections with values below 100$\mu m$) and cross sections of bullets 2 are shown on the right (cross sections with values above 100$\mu m$). \hh{Signatures are manually aligned, resulting in many of the striation marks to continuously pass from one bullet to the other. Visually, this allows for an easy assessment of these two bullet lands as a match. However, this match relies on visual inspection and can therefore not be completely objective.  The goal of this section is to eliminate the need for a visual inspection during the matching process and replace it by an automatic algorithm. This also allows for a quantification of the quality of the match.}

@@

lines <- matches %>% group_by(lineid) %>% summarise(
  meany = mean(y, na.rm=T),
  miny = min(y, na.rm=T),
  maxy = max(y, na.rm=T) + 1.5625,
  match = isMatch(lineid, type),
  type = type1[1]
)
lines <- subset(lines, lineid != 0)
@

\subsection{Horizontal Alignment}
A single cross section across the surface of the two lands in Figure~\ref{fig:manualmatch-rgl} is shown in Figure~\ref{fig:cross100}.  This figure gives the signatures of bullets 1-5 and 2-1 at a surface cross section of $x = 100$. Striation marks show up in this representations as peaks and valleys.  The individual characteristics are quite prominent and, again, suggest a match between the lands. A horizontal shift of one of the signatures (result shown in (b)) emphasizes the strong similarities. 
%
\begin{figure}[hbtp]
\centering
\begin{subfigure}[b]{.49\textwidth}\centering
\caption{Raw bullet land signatures.\label{fig:crosscut}}{%
<<crosscuts, echo=FALSE, dependson='twolands100', fig.width=7, fig.height=3, out.width='\\textwidth'>>=
qplot(y, resid, group=bullet, colour=bullet, data=lof, 
      geom="line") + 
  scale_colour_manual("", values=cols) + theme_bw() +
  theme(legend.position = c(.9, .2),
        legend.background = element_rect(fill=alpha('white', 0.4))) + 
  ylab("Residuals of loess fit")
@
    }
\end{subfigure}    
\begin{subfigure}[b]{.49\textwidth}\centering
    \caption{Aligned (by horizontal shift) signatures.\label{fig:crosscutX}}{%
<<crosscutsX, echo=FALSE, dependson='twolands100', fig.width=7, fig.height=3, out.width='\\textwidth'>>=
qplot(y+min(lof$y), resid, group=bullet, colour=bullet, data=lofY, 
      geom="line") + xlab("y") +
  scale_colour_manual("", values=cols) + theme_bw() +
  theme(legend.position = c(.9, .2),
        legend.background = element_rect(fill=alpha('white', 0.4))) + 
  ylab("Residuals of loess fit")
@
    }
\end{subfigure}
\caption{\label{fig:cross100}Signatures of bullets 1-5 and 2-1 at cross sections of $x = 100\mu m$. A horizontal shift of the values of bullet 1-5 to the right shows the similarity of the striation marks.}
\end{figure}
%
This horizontal shift is based on the cross-correlation between the two signatures: let $f(t)$ and $g(t)$ define the signature values  at $t$, where $t$ are locations between 0 $\mu m$ and about 2500 $\mu m$, 1.5625 $\mu m$ apart.
The cross-correlation between $f$ and $g$ at lag $k$ is then defined as
\[
(f * g) (k) = \sum_t f(t+k) g(t),
\]
with suitably defined limits for the summation. % (because variances of $f$ and $g$ do not change, the above equation is proportional to the correlation).

\begin{figure}[hbtp]
  \centering
<<ccf, echo=FALSE, fig.width = 7, fig.height = 3, out.width = '.5\\textwidth'>>=
lag <- ccf$lag[which.max(ccf$acf)]
qplot(x=ccf$lag, xend=ccf$lag, y = 0, yend= ccf$acf, geom="segment", colour = I("grey50")) + 
  theme_bw() + ylab("Correlation") + xlab("Lag k") +
  geom_segment(aes(x = lag, xend = lag, y = 0, yend =  max(ccf$acf)), colour="black") + 
  scale_x_continuous(breaks = c(-100, -50, lag, 0, 50, 100),
                     minor_breaks = c(-75,-25,25, 75))
@
\caption{\label{fig:ccf}Cross-correlation function between the two signatures shown in Figure~\ref{fig:crosscut} at lags between -100 and 100. At a lag of $\Sexpr{lag}$ the correlation peaks, indicating the largest amount of agreement between the signatures. Figure~\ref{fig:crosscutX} shows the lag-shifted signatures.}
\end{figure}

\subsection{Varying cross sections}
The cross section on which a comparison between signatures is based matters -- signatures from cross sections that are further apart, show more pronounced differences between the signatures. 
This poses both a caveat to matching attempts as well as an opportunity \hh{for quality control}: on the one hand we have to be aware of which cross section level was used in a matching, but on the other hand we can extract signatures from multiple cross sections of the same bullet for an initial assessment of its quality. \hh{XXX shift cross section discussion into the supplement XXX}
Both of these situations are assessed and discussed in more detail below:

%\afterpage{\clearpage}

@@

\begin{figure}[hbtp]
  \centering
<<tree:forest, echo=FALSE, fig.width=5, fig.height=5.25, out.width='.5\\textwidth', warning = FALSE>>=
qplot(pred, forest, geom="jitter", data=bstats, colour=match, 
      shape=match) + theme_bw() +
  theme(legend.position="bottom") +
  geom_vline(xintercept=0.5, colour="grey50") +
  geom_hline(yintercept=0.5, colour="grey50") + 
  xlab("Predictions from decision tree") +
  ylab("Predictions from random forest")   +
  scale_colour_brewer(palette="Paired", labels=c("KNM", "KM")) +
  scale_shape_discrete(labels=c("KNM", "KM"))
@
\caption{\label{fig:tree-forest}Scatterplot of prediction results from the tree (horizontal) and the forest (vertical). Each point corresponds to a land-to-land comparison. Known matches are indicated in blue. Matches predicted by the decision tree are on the right half of the plot, matches predicted by the random forest are in the upper half. }
\end{figure}

### 2015-12-23 imaging-supplement.Rnw
more into supplement

\end{document}

>>>>

\subsection{Assessing cross-correlation between signatures at multiple cross sections}

Figure~\ref{fig:crosscuts} shows a sequence of cross sections for bullet 1-5 (barrel 1) that are taken 50$\mu m$ apart, between 150$\mu m$  and 400$\mu m$. These are compared to a cross section at 100$\mu m$. Initially this comparison constitutes an almost perfect match between the two cross sections. However, the match quickly deteriorates with increasing distance to non-matches at a cross section of $x = 300\mu m$ or higher.  Only if the cross section is within 150$\mu m$ do we get a good visual match even when we know that the same bullet surface is being used. 
Given that we have to expect some variation in nominally the same cross section levels due to manual alignments in microscopes, we should take cross section levels into account in the automatic matching routine by evaluating matches at several cross sections. 
\begin{figure}[hbtp]
  \centering
<<crosscuts-vary, echo=FALSE, fig.width = 12, fig.height = 7, out.width = '\\textwidth', cache=TRUE, warning = FALSE>>=
paths <- file.path(datadir, dir(datadir))
crosscuts <- seq(100, 400, by = 50)
lof <- processBullets(paths[5], x = crosscuts)
lof$bullet <- paste(lof$bullet, lof$x)

reslist <- lapply(crosscuts[-1], function(cc) {
#  browser()
  b2 <- subset(lof, x %in% c(cc, 100))
  lofX <- bulletSmooth(b2)
  bAlign = bulletAlign(lofX)
  lofX <- bAlign$bullet
    b12 <- unique(b2$bullet)
  peaks1 <- get_peaks(subset(lofX, bullet==b12[1]), smoothfactor = 25)
  peaks2 <- get_peaks(subset(lofX, bullet == b12[2]), smoothfactor = 25)

#  threshold <- bulletPickThreshold(lofX, thresholds = seq(0.3, 1.5, by = 0.05))
#  lines <- striation_identify(lofX, threshold = threshold)
  peaks1$lines$bullet <- b12[1]
  peaks2$lines$bullet <- b12[2]
  lines <- striation_identifyXXX(peaks1$lines, peaks2$lines)

  maxCMS <- maxCMS(lines$match==TRUE)
  list(maxCMS = maxCMS, ccf = bAlign$ccf, lines=lines, bullets=lofX)
})

ccfs <- sapply(reslist, function(res) res$ccf)

lop <- lapply(reslist, function(res) {
ggplot() +
  theme_bw() + 
  geom_rect(aes(xmin=xmin, xmax=xmax, fill=factor(type)), show.legend=FALSE, ymin=-6, ymax=5, data=res$lines, alpha=0.2) +
  geom_line(aes(x = y, y = l30, linetype=bullet),   data = res$bullets, alpha=0.6) +
  scale_colour_brewer("", palette="Set1", na.value=alpha("grey50", alpha=0.5)) +
  scale_linetype_discrete("") +
  scale_fill_brewer("", palette="Set2", na.value=alpha("grey50", alpha=0.5)) +
  theme(legend.position = c(1,1.2), legend.justification=c(1,1),
        legend.background = element_rect(fill=alpha('white', 0.4))) + 
  ylim(c(-6,6)) +
  geom_text(aes(x = meany), y= -5.5, label= "x", data = subset(res$lines, !match)) +
  geom_text(aes(x = meany), y= -5.5, label= "o", data = subset(res$lines, match)) +
    ylab("") + xlab("")
})

grid.arrange(lop[[1]], lop[[2]], lop[[3]], lop[[4]], lop[[5]], lop[[6]],
             ncol = 2)
@
\caption{\label{fig:crosscuts}Overview of the variations in the surface along different cross sections. Cross section at $x = 100$ is compared to cross sections every 50$\mu m$. With every step away from the original cross section site, the number of differences between the cross sections increases, and the number of maximum CMS decreases from initially 14 to  3 or fewer at $x = 300\mu m$ and higher. }
\end{figure}
\end{document}
